---
title: "Hyperparameter Estimation with openEBGM"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hyperparameter Estimation with openEBGM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Background

DuMouchel used an "Empirical Bayes" (EB) method. That is, the data are a driving
force behind the choice of the prior distribution (in contrast to typical
Bayesian methods, where the prior is chosen without regard to the actual data).
One outcome of this is a known posterior distribution that relies on a set of
hyperparameters derived from the prior distribution. These hyperparameters are
estimated by their most likely value in the context of Empirical Bayesian
methods. One process by which this can occur involves maximizing the likelihood
function of the marginal distributions of the counts (or in our case, minimizing
the negative log-likelihood function).

### Optimization

Global optimization is a broad field. There are many existing R packages with
minimization routines which may be used in the estimation of these 
hyperparameters. The hyperparameter estimation functions offered by *openEBGM*
utilize the following:

* Optimization using PORT routines
* Newton-type algorithm
* A quasi-Newton method (also known as a variable metric method)
* A version of the Expectation-Maximization (EM) Algorithm known as the
Expectation/Conditional Maximization (ECM) Algorithm (*1*)

*openEBGM*'s hyperparameter estimation functions use local optimization
algorithms. The Newton-like approaches use algorithms implemented in R's *stats*
package and allow the user to choose multiple starting points to improve the
chances of finding a global optimum. The user is encouraged to explore a variety
of optimization approaches because the accuracy of a global optimization result
is extremely difficult to verify and other approaches might work better in some
cases.

### References

1. Meng X-L, Rubin D (1993). "Maximum likelihood estimation via the ECM
algorithm: A general framework." *Biometrika*, 80(2), 267-278.
*`https://doi.org/10.1093/biomet/80.2.267`*

1. DuMouchel W, Pregibon D (2001). "Empirical Bayes Screening for
Multi-item Associations." In *Proceedings of the Seventh ACM SIGKDD
International Conference on Knowledge Discovery and Data Mining*, KDD '01,
pp. 67-76. ACM, New York, NY, USA. ISBN 1-58113-391-X.
*`https://doi.org/10.1145/502512.502526`*

1. Millar, Russell B (2011). "Maximum Likelihood Estimation and Inference."
*John Wiley & Sons, Ltd*, 111-112.
*`https://doi.org/10.1002/9780470094846`*

--------

# Estimating the Hyperparameters

## Data Squashing

DuMouchel & Pregibon (*2*) discuss a methodology they call "data squashing", 
which "compacts" the dataset to reduce the amount of computation needed to
estimate the hyperparameters. *openEBGM* provides an implementation for data 
squashing.

The actual counts ($N$) and expected counts ($E$) are used to estimate the 
hyperparameters of the prior distribution. A large contingency table will have 
many cells, resulting in computational difficulties for the optimization 
routines needed for estimation. Data squashing (DuMouchel et al., 2001) 
transforms a set of 2-dimensional points to a smaller number of 3-dimensional 
points. The idea is to reduce a large number of points $(N, E)$ to a smaller 
number of points $(N_k, E_k, W_k)$, where $k = 1,...,M$ and $W_k$ is the weight
of the $k^{th}$ "superpoint". To minimize information loss, only points close to
each other should be squashed.

For a given $N$, `squashData()` combines points with similar $E$s into bins
using a specified bin size and uses the average $E$ within each bin as the $E$
for that bin's "superpoint". The new superpoints are weighted by bin size. For
example, the points (1, 1.1) and (1, 1.3) could be squashed to the superpoint
(1, 1.2, 2).

An example is given below using unstratified expected counts:

```{r dataSquash example}
library(openEBGM)
data(caers)

processed <- processRaw(caers)
processed[1:4, 1:4]

squashed <- squashData(processed) #Using defaults
head(squashed)

nrow(processed)
nrow(squashed)
```

As shown above, the squashed data set has
`r round(nrow(squashed)/nrow(processed)*100, 2)`% of the observations as the
full dataset. Using this squashed data, we can then estimate the hyperparameters
in a far more efficient manner.

`squashData()` can be used iteratively for each count ($N$):

```{r iterative squashing example}
squash1 <- squashData(processed)
squash2 <- squashData(squash1, count = 2, bin_size = 8)
```

Or `autoSquash()` can be used to squash all counts at once:

```{r autoSquash example}
squash3 <- autoSquash(processed)
ftable(squash3[, c("N", "weight")])
```

## Likelihood Functions

As previously mentioned, the hyperparameters are estimated by minimizing the
negative log-likelihood function. There are actually 4 different functions, 
depending on the use of data squashing and zero counts. All 4 functions, 
however, are based on the marginal distribution of the counts, which are 
mixtures of two negative binomial distributions. The hyperparameters are denoted
by the vector $\theta=(\alpha_1,\beta_1,\alpha_2,\beta_2,P)$, where $P$ is 
the mixture fraction.

The most commonly used likelihood function is `negLLsquash()`, which is used 
when squashing data and not using zero counts. `negLLsquash()` is not called 
directly, but rather by some optimization function:

```{r negLLsquash example, warning = FALSE}
theta_init <- c(alpha1 = 0.2, beta1 = 0.1, alpha2 = 2, beta2 = 4, p = 1/3)
stats::nlm(negLLsquash, p = theta_init,
           ni = squashed$N, ei = squashed$E, wi = squashed$weight, N_star = 1)
```

The `N_star` argument allows the user to choose the smallest value of $N$ used 
for hyperparameter estimation. The user must be careful to match `N_star` with 
the actual minimum count in `ni`. If the user wishes to use a larger value for 
`N_star`, the vectors supplied to arguments `ni`, `ei`, and `wi` must be
filtered. Here, we are using all counts except zeroes, so no filtering is 
needed. In general, `N_star = 1` should be used whenever practical.

Note that you will likely receive warning messages from the optimization 
function--use your own judgment to determine whether or not they would likely
indicate real problems.

The other likelihood functions are `negLL()`, `negLLzero()`, and
`negLLzeroSquash()`. Make sure to use the appropriate function for your choice
of data squashing and use of zero counts.

## Hyperparameter Estimation Functions

Hyperparameters can be calculated by exploring the parameter space of the 
likelihood function using either the full data set of $N$s and $E$s or the 
squashed set. The methodology implemented by this package essentially maximizes 
the likelihood function (or more specifically, minimizes the negative 
log-likelihood function). Starting points must be chosen to begin the 
exploration. DuMouchel (1999, 2001) provides a "lightly justified" set of 
initial hyperparameters. However, *openEBGM*'s functions support a large set of
starting choices to help reach convergence and reduce the chance of false
convergence. We begin by defining some starting points:

```{r}
theta_init <- data.frame(alpha1 = c(0.2, 0.1, 0.3),
                         beta1  = c(0.1, 0.2, 0.5),
                         alpha2 = c(2,   10,  6),
                         beta2  = c(4,   10,  6),
                         p      = c(1/3, 0.2, 0.5)
)
```

###`autoHyper()`

Now that the initial guesses for the hyperparameters have been defined, the 
function `autoHyper()` can be used to determine the actual hyperparameter 
estimates. `autoHyper()` performs a "verification check" by requiring the 
optimization routine to converge at least twice within the bounds of the 
parameter space. The estimate corresponding to the smallest negative 
log-likelihood is chosen as a tentative result. By default, this estimate must 
be similar to at least one other convergent solution. If the algorithm fails to
consistently converge, an error message is returned.

```{r autoHyper example, warning = FALSE}
system.time(
hyper_estimates_full <- autoHyper(data = processed, theta_init = theta_init, 
                                  squashed = FALSE)
)

squashed2 <- squashData(squashed, count = 2, bin_size = 10, keep_pts = 50)
system.time(
hyper_estimates_squashed <- autoHyper(data = squashed2, theta_init = theta_init)
)

hyper_estimates_full

hyper_estimates_squashed
```

As seen above, the process is much faster when utilizing the squashed data, with
estimates that are nearly identical. Of course, the amount of efficiency
increase depends on the parameter values used in the call to `squashData()` and
the size of the original data set. Another factor affecting run time is the
number of starting points used. In general, you should use five or fewer
starting points and limit the number of data points to a maximum of about 20,000
if possible. Notice that we squashed the same data again, which is allowed if we
use a different value for `count` each time.

`autoHyper()` utilizes multiple minimization techniques to verify convergence. 
It first attempts the `stats::nlminb()` function, which implements a 
quasi-Newton unconstrained optimization technique using PORT routines. If 
`nlminb()` fails to consistently converge, `autoHyper()` next attempts 
`stats::nlm()`, which is a non-linear maximization algorithm based on the Newton
method. Finally, if the first two approaches fail, a quasi-Newton method (also 
known as a variable metric algorithm) is used, which "...uses function values 
and gradients to build up a picture of the surface to be optimized." (source: R
documentation for stats::optim) This routine is implemented by the 
`stats::optim()` function with the `method = BFGS` argument.

`autoHyper()` can now return standard errors and confidence intervals for the
hyperparameter estimates:

```{r autoHyper example with CIs, warning = FALSE}
autoHyper(squashed2, theta_init = theta_init, conf_ints = TRUE)$conf_int
```

###`exploreHypers()`

`autoHyper()` is a semi-automated (but imperfect) approach to hyperparameter 
estimation. The user is encouraged to explore various optimization approaches as
no single approach will always work (the functions in *openEBGM* are not the
only optimization functions available in R). One way to explore is with
*openEBGM*'s `exploreHypers()` function, which is actually called by
`autoHyper()` behind-the-scenes. `exploreHypers()` can now also return standard
errors for the hyperparameter estimates:

```{r hyperparameter estimation example, warning = FALSE}
exploreHypers(data = squashed2, theta_init = theta_init, std_errors = TRUE)
```

`exploreHypers()` offers three gradient-based optimization methods and is 
basically just a wrapper around commonly used functions from the *stats* package
mentioned earlier: `nlminb()` (the default), `nlm()`, & `optim()`.

###`hyperEM()`

`hyperEM()` implements a version of the EM algorithm referred to by Meng & Rubin
(*1*) as the Expectation/Conditional Maximization (ECM) algorithm. `hyperEM()`
uses a single starting point to find a local maximum likelihood estimate for
$\theta$ either by finding roots of the score functions (partial derivatives of
the log-likelihood function) or by using `stats::nlminb()` to directly minimize
the negative log-likelihood function. The method described by Millar (*3*) is
used to accelerate the estimate of $\theta$ every 100 iterations of the
algorithm.

```{r hyperEM example, warning = FALSE}
data(caers)
proc <- processRaw(caers)
squashed <- squashData(proc, bin_size = 100, keep_pts = 0)
squashed <- squashData(squashed, count = 2, bin_size = 12)
hyperEM_ests <- hyperEM(squashed, theta_init_vec = c(1, 1, 2, 2, .3),
                        conf_ints = TRUE, track = TRUE)
str(hyperEM_ests)
```

Setting `track = TRUE` tracks the log-likelihood and hyperparameter estimates at
each iteration, which we can plot to study the behavior of the algorithm:

```{r hyperEM plotting example, warning = FALSE, fig.width = 7, fig.height = 10}
library(ggplot2)
library(tidyr)
pdat <- gather(hyperEM_ests$tracking, key = "metric", value = "value", logL:P)
pdat$metric <- factor(pdat$metric, levels = unique(pdat$metric), ordered = TRUE)
ggplot(pdat, aes(x = iter, y = value)) +
  geom_line(size = 1.1, col = "blue") +
  facet_grid(metric ~ ., scales = "free") +
  ggtitle("Convergence Assessment",
          subtitle = "Dashed red line indicates accelerated estimate") +
  labs(x = "Iteration Count", y = "Estimate") +
  geom_vline(xintercept = c(100, 200), size = 1, linetype = 2, col = "red")
```

### *DEoptim* package

Other packages, such as
[DEoptim](https://cran.r-project.org/package=DEoptim){target="_blank"},
can alternatively be used to estimate the hyperparameters:

```{r DEoptim example, warning = FALSE, message = FALSE}
library(DEoptim)
set.seed(123456)
theta_hat <- DEoptim(negLLsquash,
                     lower = c(rep(1e-05, 4), .001),
                     upper = c(rep(5, 4), 1 - .001),
                     control = DEoptim.control(
                       itermax = 2000,
                       reltol  = 1e-04,
                       steptol = 200,
                       NP      = 100,
                       CR      = 0.85,
                       F       = 0.75,
                       trace   = 25
                     ),
                     ni = squashed$N, ei = squashed$E, wi = squashed$weight
)
(theta_hat <- as.numeric(theta_hat$optim$bestmem))
```

Although the user is encouraged to explore various approaches for maximum 
likelihood hyperparameter estimation, `autoHyper()` will often give
reasonable results with minimal effort. Once the hyperparameters have been
estimated, they can be used in the calculation of the $EBGM$ and quantile scores
by applying them to the posterior distribution. This process can be found in the
[*Empirical Bayes Metrics with openEBGM*](https://cran.r-project.org/package=openEBGM/vignettes/x4_posteriorCalculationVignette.html){target="_blank"}
vignette.
