---
title: "Introduction to openEBGM"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to openEBGM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Background

William DuMouchel (*1*, *2*) created an empirical Bayes (EB) data mining
approach for finding "interestingly large" counts in contingency tables.
DuMouchel's approach works well even when most of the counts are zero or one
(i.e., a sparse table). The benefit of DuMouchel's model over simpler approaches
such as the *relative reporting ratio*, $RR$, is that Bayesian shrinkage
corrects for the high variability in $RR$ that results from small counts.

The rows and columns of the table represent levels of two different variables,
such as food or drug products and adverse events:

|                | Headache  |Nausea |Vomiting | ...   |
| :-------------:|:---------:|:-----:|:-------:|:-----:|
| **Product A**  | 0         |1      |0        |**...**|
| **Product B**  | 4         |0      |0        |**...**|
| **Product C**  | 1         |0      |9        |**...**|
| **...**        | **...**   |**...**|**...**  |**...**|

The relative reporting ratio is calculated as $RR=\frac{N}{E}$, where $N$ is the
actual count for a cell in the table and $E$ is the expected count under the
assumption of independence between rows and columns. When $RR = 1$, you observe
the exact count you would expect to observe if no association exists between the
two variables. When $RR > 1$, you observe a larger count than expected. This
approach works well for large counts; however, small counts cause $RR$ to become
quite unstable. For instance, an actual count of $N = 1$ with an expected count
of $E = 0.01$ gives us $RR = 100$ -- which seems large -- but a single event
could easily occur simply by chance. The EB approach shrinks large $RR$s that
result from small counts to a value much closer to the "null hypothesis" value
of 1. The shrinkage is smaller for larger counts and negligible for very large
counts. Shrinkage gives results that are more stable than the simple $RR$
measurement.

DuMouchel's model uses a Poisson($\mu_{ij}$) likelihood (i.e. data distribution)
for the actual cell counts, $N_{ij}$, in row *i* and column *j*. The expected
cell counts, $E_{ij}$, are treated as constants. We are interested in the ratio
$\lambda_{ij}=\frac{\mu_{ij}}{E_{ij}}$, which is analagous to $RR=\frac{N}{E}$.
The prior on $\lambda$ is a mixture of two gamma distributions, resulting in a
posterior distribution for $\lambda$ which is a mixture of two gamma
distributions. Hence, the model is sometimes referred to as the **Gamma-Poisson
Shrinker (GPS)** model. The posterior distribution of $\lambda$ can be thought
of as a Bayesian representation of $RR$. Summary statsistics from the posterior
distribution are used as attenuated versions of $RR$.

Each cell in the contingency table will have its own posterior distribution
determined both by that cell's actual and expected counts (the data) and by the
distribution of actual and expected counts in the entire table (the prior).
Often, the Empirical Bayes Geometric Mean $(EBGM)$ of the posterior distribution
is used in place of $RR$. Alternatively, the more conservative percentiles (5th,
10th, etc.) can be used. The percentiles can also be used to construct Bayesian
credible intervals. Similar to $RR$, an $EBGM$ (or lower percentile) much bigger
than 1 represents an actual count much bigger than expected. Such cases might
represent signals of interest, and the product/event pair can be further
examined by subject matter experts to determine if the association might
actually be causal in nature.

An extension of the GPS model, the Multi-Item Gamma-Poisson Shrinker (MGPS) 
model (2001), is currently being used by the U.S. Food and Drug Administration 
(FDA) to find higher-than-expected reporting of adverse events associated with 
food, drugs, etc. For instance, FDA's Center for Food Safety and Applied 
Nutrition (CFSAN) uses the MGPS model to mine data from the CFSAN Adverse Events
Reporting System (CAERS): <https://www.fda.gov/Food/ComplianceEnforcement>{target="_blank"}. (The 
variables forming the rows and columns of the contingency table are *product*
and *adverse event*.) MGPS allows for product interactions, unlike the GPS model
implemented in *openEBGM* (*3*), which can only use individual product-event
pairs.

## Purpose

The *openEBGM* package implements DuMouchel's approach with some small
differences. For example, the expected counts are calculated by counting unique
"transactions" (*2*) in each row and column, not actual marginal totals. In the
CAERS data, a unique report is a transaction. In some applications, a single
transaction could occur in several cells. For instance, a single CAERS report
might mention multiple products and/or adverse events. Using simple marginal
totals would then count a single report multiple times.

This document teaches you how to prepare your data for use by *openEBGM*'s 
functions. Other vignettes give explanations and examples of more advanced 
topics:

* **Raw data processing:** Process your data to find counts and simple
disproportionality measures.

* **Hyperparameter estimation:** Estimate the hyperparameters of the prior 
distribution.

* **Empirical Bayes metrics:** This is the ultimate goal. Calculate Empirical
Bayes metrics ($EBGM$ and quantile scores) based on the posterior distribution.

* **Object-oriented features:** Create objects of a special class (*openEBGM*)
to use with generic functions such as `plot()`.

### References

1. DuMouchel W (1999). "Bayesian Data Mining in Large Frequency
Tables, With an Application to the FDA Spontaneous Reporting System."
*The American Statistician*, 53(3), 177-190. 
*`https://doi.org/10.1080/00031305.1999.10474456`*

1. DuMouchel W, Pregibon D (2001). "Empirical Bayes Screening for
Multi-item Associations." In *Proceedings of the Seventh ACM SIGKDD
International Conference on Knowledge Discovery and Data Mining*, KDD '01,
pp. 67-76. ACM, New York, NY, USA. ISBN 1-58113-391-X.
*`https://doi.org/10.1145/502512.502526`*

1. Canida T, Ihrie J (2017). "openEBGM: An R Implementation of the Gamma-Poisson
Shrinker Data Mining Model." *The R Journal*, 9(2), 499-519.
*`https://journal.r-project.org/archive/2017/RJ-2017-063/index.html`*


---------

# Preparing Your Data

*openEBGM* requires the input data to be a data frame of a particular form.

## Data form

Data must be in tidy format (one column per variable and one observation per
row). The columns can be of type factor, character, integer, or numeric. Missing
values are not allowed - either replace NAs and empty strings with appropriate
values or remove them from the data.


## Column names

The input data frame must contain certain column names: *var1*, *var2*, and 
*id*.  *var1* and *var2* are simply the row and column variables of the 
contingency table. The identifier (*id*) column allows *openEBGM* to properly 
handle marginal totals (for instance, this would be the report identifier 
variable in the aformentioned CAERS data). If the cells of the table actually
represent mutually exclusive "events of interest", the user can create a column
of unique sequential identifiers with `df$id <- 1:nrow(df)`.

Stratification can help reduce the effects of confounding variables. If
stratification is used, any column whose name contains the substring *'strat'*
(case sensitive) will be treated as a stratification variable. If a continuous
variable such as age is used for stratification, remember to categorize the
variable.

Other columns are allowed, but will be ignored.

## CAERS example

Here is a small subset of raw data from the publicly available [CAERS data](https://www.fda.gov/Food/ComplianceEnforcement){target="_blank"} described
above:

```{r Load data}
library(openEBGM)
data(caers_raw)
head(caers_raw, 4)
```

Only one product name is given per row, but we need to separate the adverse 
events into different rows:

```{r One adverse event per row}
dat <- tidyr::separate_rows(caers_raw, SYM_One.Row.Coded.Symptoms, sep = ", ")
dat[1:4, c("RA_Report..", "PRI_Reported.Brand.Product.Name", 
           "SYM_One.Row.Coded.Symptoms")]
```

Next we need to change column names:

```{r Rename columns}
dat$id   <- dat$RA_Report..
dat$var1 <- dat$PRI_Reported.Brand.Product.Name
dat$var2 <- dat$SYM_One.Row.Coded.Symptoms
```

Suppose we want to use gender and age as stratification variables:

```{r Rename gender column}
dat$strat_gender <- dat$CI_Gender
table(dat$strat_gender, useNA = "always")
```

*Age* is a continuous variable, so we need to create categories:

```{r Categorize continuous variables}
dat$age_yrs <-
  ifelse(dat$CI_Age.Unit == "Day(s)", dat$CI_Age.at.Adverse.Event / 365,
  ifelse(dat$CI_Age.Unit == "Decade(s)", dat$CI_Age.at.Adverse.Event * 10,
  ifelse(dat$CI_Age.Unit == "Month(s)", dat$CI_Age.at.Adverse.Event / 12,
  ifelse(dat$CI_Age.Unit == "Week(s)", dat$CI_Age.at.Adverse.Event / 52,
  ifelse(dat$CI_Age.Unit == "Year(s)", dat$CI_Age.at.Adverse.Event,
         NA)))))
dat$strat_age <- ifelse(is.na(dat$age_yrs), "unknown",
                 ifelse(dat$age_yrs < 18, "under_18",
                        "18_plus"))
table(dat$strat_age, useNA = "always")
```

Now we have the data in the proper form:

```{r Display prepared data}
vars <- c("id", "var1", "var2", "strat_gender", "strat_age")
dat[1:5, vars]
```

Next, the [*Processing Raw Data with openEBGM*](https://cran.r-project.org/package=openEBGM/vignettes/x2_rawDataProcessingVignette.html){target="_blank"}
vignette will demonstrate how to use data in this general form to find counts
and simple disproportionality measures--$RR$ and $PRR$ (proportional reporting
ratio).
